---
title: "Assignment_1"
output: html_notebook
---

```{r}
library(tidyr)
library(ggplot2)
library(MASS)
library(ggdensity)
library(data.table)
library(dplyr)

```

```{r}
#Load data
data_PJM <- fread('PJM Data.csv')
data_Chicago <- fread('Data Chicago.csv')
data_Milwaukee <- fread('Data Milwaukee.csv')
data_CorpusChristi<- fread('Data Corpus Christi.csv')
```

```{r}
#Exploring data
names(data_PJM)
```

```{r}
#Units of observation
class(data_PJM$datetime_beginning_utc)
```

```{r}
#Identify initial date
min(data_PJM$date)
```

```{r}
#Unique price_node names
head(unique(data_PJM$pnode_name),5)
```

```{r}
#number of different pricing nodes
length(unique(data_PJM$pnode_name))
```

```{r}
#Interesting fields to compares
data_PJM[, .(date, congestion_price_rt)]
```

```{r}
library(dplyr)
#dataset of all observation on a holiday
holiday_PJM <- data_PJM[isHoliday == 'TRUE']
```

```{r}
avg_lmp_chicago_holiday = holiday_PJM[month == 'Dec' &
pnode_name == 'CHICAGO GEN HUB' &
am_pm == 'AM',
.(average_lmp_rt = mean(total_lmp_rt)), by = .(hour, year)] %>%
.[order(year)]

```

```{r}
#Plot to visulize data
library(ggplot2)

ggplot(avg_lmp_chicago_holiday, aes(x = hour, y = average_lmp_rt, color = factor(year))) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Avg Hourly LMP on Holidays (December, AM) - CHICAGO GEN HUB",
       x = "Hour of Day", y = "Average LMP (Real-Time)",
       color = "Year") +
  theme_minimal()
```

```{r}

#2022 shows the highest average locational marginal price at the Chicago gen hub node on holidays
#The data is being averaged over total_lmp_rt. This could be based on the winter storm chicago may have #been having in the holiday months. We can cross check the data by running a regression against #temperature, and seeing if there is any predictive power between weather and average_lmp_rt

```

```{r}
#Introduction to plotting
```

```{r}
#creating data_CGH
data_CGH = data_PJM[pnode_name == 'CHICAGO GEN HUB']
#Number of observations refect the number of rows
nrow(data_CGH)
```

```{r}
data_CGH$day_of_week <- factor(data_CGH$day_of_week, 
levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

ggplot(data_CGH[order(day_of_week)], aes(x=day_of_week,y=total_lmp_rt, color =
    factor(year))) +
    stat_summary() +
    labs(color = '', x = '', y = 'Total RT LMP ($/MWh)') +
    coord_cartesian(ylim = c(0, 80)) +
    scale_y_continuous(expand = c(0,0)) +
    theme_classic() +
    theme(legend.position = 'top')
```

```{r}

#We can see the total lmp_rt per a given day across 2021-2023. We can see that consistently had the #highest locational margin price throughout the week

```

```{r}
#remove color = factor(year)
data_CGH$day_of_week <- factor(data_CGH$day_of_week, levels =
c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday",
"Sunday"))
ggplot(data_CGH[order(day_of_week)], aes(x=day_of_week,y=total_lmp_rt)) +
stat_summary() +
labs(color = '', x = '', y = 'Total RT LMP ($/MWh)') +
coord_cartesian(ylim = c(0, 80)) +
scale_y_continuous(expand = c(0,0)) +
theme_classic() +
theme(legend.position = 'top')
```

```{r}
#When color = factor(year) is removed it looks like we are seeing the average across all 3 years

```

```{r}
# By month
data_CGH$month_name <- factor(data_CGH$month, levels = 
  c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
    "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

ggplot(data_CGH[order(month)], aes(x = month_name, y=total_lmp_rt, color =
factor(year))) +
stat_summary() +
labs(color = '', x = '', y = 'Total RT LMP ($/MWh)') +
coord_cartesian(ylim = c(0, 80)) +
scale_y_continuous(expand = c(0,0)) +
theme_classic() +
theme(legend.position = 'top')
```

```{r}
#by hour
ggplot(data_CGH, aes(x=hour, y=total_lmp_rt, 
        color = factor(year))) +
        stat_summary() +
        labs(color = '', x = '', y = 'Total RT LMP ($/MWh)') +
        coord_cartesian(ylim = c(0, 80)) +
        scale_y_continuous(expand = c(0,0)) +
        theme_classic() +
        theme(legend.position = 'top')

```

```{r}
avg_lmp_chicago <- data_CGH[, .(average_lmp_rt = mean(total_lmp_rt)), 
                            by = .(month, year)]

avg_lmp_chicago$month_name <- factor(avg_lmp_chicago$month, levels = 
  c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
    "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

ggplot(avg_lmp_chicago[order(month_name)], 
       aes(x = month_name,
           y = average_lmp_rt,
           color = factor(year))) +
      stat_summary() +
      labs(color = '',x = '', y = 'Average RT LMP ($/MWh)') +
      coord_cartesian(ylim = c(0,80)) +
      theme_classic() + 
      theme(legend.position = 'top')
```

```{r}
#Chicago Summer '22 saw above higher temperatures #(https://www.weather.gov/lot/Summer_August_2022_Climate_Summaries)
#Also the russia-ukraine war could have been a factor that impact crude oil price and electricity is #generated from oil, among other forms

```

```{r}
#pivoting data total_lmp_rt, and total_lmp_da
library(tidyr)
library(ggplot2)
data_CGH_long <- data_CGH[, .(year, month, hour, day_of_week, total_lmp_rt,total_lmp_da)] %>%
  pivot_longer(cols = !c(year, month, hour, day_of_week),
               names_to = 'lmp_type',
               values_to = 'values')
```

```{r}
#This goes and makes two columns into rows by creating a new column lmp type and putting
#total_lmp_rt and total_lmp_da as values
```

```{r}
head(data_CGH_long)
```

```{r}
setDT(data_CGH_long) #go from object to data table

ggplot(data_CGH_long, aes(x=day_of_week,y=values, color = factor(lmp_type))) +
        stat_summary() +
        labs(color = '', x = '', y = 'LMP ($/MWh)') +
        coord_cartesian(ylim = c(0, 100)) +
        scale_y_continuous(expand = c(0,0)) +
        theme_classic() +
        theme(legend.position = 'top',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
        facet_wrap(~year)
```

```{r}
data_CGH_long$month_name <- factor(data_CGH_long$month, levels = 
  c("Jan", "Feb", "Mar", "Apr", "May", "Jun",
    "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
ggplot(data_CGH_long, aes(x=month_name,y=values, color = factor(lmp_type))) +
        stat_summary() +
        labs(color = '', x = '', y = 'LMP ($/MWh)') +
        coord_cartesian(ylim = c(0, 100)) +
        scale_y_continuous(expand = c(0,0)) +
        theme_classic() +
        theme(legend.position = 'top',
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
        facet_wrap(~year)
```

```{r}
CGH_summary_daily <- data_CGH[, .(mean_rt_lmp = mean(total_lmp_rt),
  mean_da_lmp = mean(total_lmp_da)),
  by = .(day, month,year)]
ggplot(CGH_summary_daily, aes(mean_rt_lmp, color = factor(year))) +
    stat_ecdf() +
    labs(color = '', x = 'Average RT LMP ($/MWh)',
    y = 'Empirical Cumulative Distribution Function') +
    theme_classic() +
    theme(legend.position = 'top')
```
```{r}
CGH_summary_daily <- data_CGH[, .(vol_lmp_rt = sd(total_lmp_rt),
  vol_lmp_da = sd(total_lmp_da)),
  by = .(day, month,year)]
ggplot(CGH_summary_daily, aes(vol_lmp_rt, color = factor(year))) +
    stat_ecdf() +
    labs(color = '', x = 'Volatility RT LMP ($/MWh)',
    y = 'Empirical Cumulative Distribution Function') +
    theme_classic() +
    theme(legend.position = 'top')
```
```{r}
vol_daily_CGH <- data_CGH[, .(vol_spread = sd(spread, na.rm = TRUE)),
by = .(day, month, year)]
  ggplot(vol_daily_CGH[month == 'Aug'], aes(x=vol_spread, color = factor(year))) +
  geom_density() +
  labs(x = 'Standard Deviation of RT LMP - DA LMP ($/MWh)', y = 'Probability Density
  Function', color = '') +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(legend.position = 'top')

```
```{r}
#We are only looking at August for each year. The vol spread is calculated as RT- DA. 2022 has the wides distribution indicating more volatility
```


```{r}
#Comparing to the eastern Hub

#1 creating datatable
data_easternHub <- data_PJM[pnode_name == 'EASTERN HUB']
#2 calculating the vol spread
vol_daily_EH <- data_easternHub[, .(vol_spread = sd(spread, na.rm = TRUE)),
by = .(day, month, year)]
#3 Plotting
  ggplot(vol_daily_EH[month == 'Aug'], aes(x=vol_spread, color = factor(year))) +
  geom_density() +
  labs(x = 'Standard Deviation of RT LMP - DA LMP ($/MWh)', y = 'Probability Density
  Function', color = '') +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0)) +
  theme_classic() +
  theme(legend.position = 'top')



```
```{r}
#We see tighter spreads, and lower volatility for the Eastern Hub
```


```{r}
#Plot the hourly by day for the month of April 2023 for Chicago GEN hum on RT LMP

# Filter April 2023
april_data <- data_CGH[month == 'Apr' & year == 2023]


ggplot(april_data, aes(x = hour, y = total_lmp_rt)) +
  geom_col(color = "red") +
  labs(title = "Hourly RT LMP for Chicago Gen Hub (April 2023)",
       x = "Hour", y = "RT LMP ($/MWh)") +
  theme_classic()
```
```{r}
#Negative LMP prices indicate that there was a greater supply than demand of energy. Since we had a surplus of energy and no place to store the cost of carry pushed prices negative. 

```

```{r}
# Filter May 2023
library(dplyr)

may_data <- data_CGH[month == 'May' & year == 2023]


ggplot(may_data, aes(x = hour, y = total_lmp_rt)) +
  geom_col(color = "blue") +
  labs(title = "Hourly RT LMP for Chicago Gen Hub (May 2023)",
       x = "Hour", y = "RT LMP ($/MWh)") +
  theme_classic()
```
```{r}
# Filter June 2023
june_data <- data_CGH[month == 'Jun' & year == 2023]


ggplot(june_data, aes(x = hour, y = total_lmp_rt)) +
  geom_col(color = "green") +
  labs(title = "Hourly RT LMP for Chicago Gen Hub (April 2023)",
       x = "Hour", y = "RT LMP ($/MWh)") +
  theme_classic()
```

```{r}
library(dplyr)
library(purrr)
mapping <- list(
  "Jan" = "Winter",
  "Feb" = "Winter",
  "Mar" = "Spring", 
  "Apr" = "Spring",
  "May" = "Spring",
  "Jun" = "Summer", 
  "Jul" = "Summer",
  "Aug" = "Summer",
  "Sep" = "Fall", 
  "Oct" = "Fall",
  "Nov" = "Fall",
  "Dec" = "Winter"
)

# Apply the mapping
data_CGH <- data_CGH %>%
  mutate(
    season = map_chr(month, ~ mapping[[.x]])
  )
```

```{r}

# Filter April 2023

library(dplyr)
library(ggplot2)

library(dplyr)
library(ggplot2)

# Filter data for 2023
data_2023_cgh <- data_CGH %>% filter(year == 2023)

# Split data by season
season_list <- split(data_2023_cgh, data_2023_cgh$season)

plot_season_lmp <- function(df, season_name) {
  ggplot(df, aes(x = hour, y = total_lmp_rt)) +
    geom_col(color = "orange") +
    labs(title = paste(season_name, "RT LMP (2023)"),
         x = "Hour", y = "RT LMP ($/MWh)") +
    theme_classic()
}

season_plots <- lapply(names(season_list), function(season_name) {
  plot_season_lmp(season_list[[season_name]], season_name)
})

for (p in season_plots) print(p)

```




```{r}
#based on the above graph it looks like the highest is around 8:00 PM and the lowest is around 4:00 AM for april 2023. This makes sense as the energy consumption levels when everyone is asleep will be low vs when everyone is awake and home.  Though when looking across all seasons, other than spring we don't see prices go negative.
```

```{r}
#daily bocplots of RT-DA LMP Spread for 2021
ggplot(vol_daily_CGH[year == 2021], aes(factor(month, levels = month.abb),
vol_spread,
color = factor(month, levels = month.abb))) +
geom_boxplot() +
labs(x = '', y = 'Standard Deviation of RT LMP - Lagged DA LMP ($/MWh)') +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1)) +
guides(color ='none') +
theme_classic() +
theme(legend.position = 'top')
```
```{r}
#2022
ggplot(vol_daily_CGH[year == 2022], aes(factor(month, levels = month.abb),
vol_spread,
color = factor(month, levels = month.abb))) +
geom_boxplot() +
labs(x = '', y = 'Standard Deviation of RT LMP - Lagged DA LMP ($/MWh)') +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1)) +
guides(color ='none') +
theme_classic() +
theme(legend.position = 'top')
```
```{r}
#2023
ggplot(vol_daily_CGH[year == 2023], aes(factor(month, levels = month.abb),
vol_spread,
color = factor(month, levels = month.abb))) +
geom_boxplot() +
labs(x = '', y = 'Standard Deviation of RT LMP - Lagged DA LMP ($/MWh)') +
theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1)) +
guides(color ='none') +
theme_classic() +
theme(legend.position = 'top')
```
```{r}
library(lubridate)

# Assuming 'vol_daily_CGH' has columns: year, month, day, vol_spread
# If 'month' is a character like "Jan", convert it to numeric
vol_daily_CGH_wide <- vol_daily_CGH %>%
  mutate(
    month_num = match(month, month.abb),  # Converts "Jan" to 1, "Feb" to 2, etc.
    date = make_date(year, month_num, day)
  ) %>%
  select(date, vol_spread)  # Keep only the columns you want
```

```{r}
vol_daily_CGH_wide %>%
  filter(vol_spread == max(vol_spread, na.rm = TRUE))
```
```{r}
# 2022-12-2023 is the date with the most volatility
```

```{r}
library(dplyr)
library(tidyr)
library(data.table)
library(ggplot2)

daily_temp_Chicago <- data_Chicago[, .(date, month, year, 
                                       avg_temperature_degrees_fahrenheit, 
                                       avg_cooling_degree_days, 
                                       avg_heating_degree_days)] %>%
  unique() %>%
  pivot_longer(
    cols = starts_with('avg'),
    names_to = 'temp_series',
    values_to = 'degrees'
  )

setDT(daily_temp_Chicago)

ggplot(daily_temp_Chicago, aes(date, degrees, color = factor(temp_series))) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c('red','blue','black')) +
  geom_hline(yintercept = 65, color = 'forestgreen') +
  labs(
    x = '',
    color = '',
    y = 'Temperature (Degrees Fahrenheit)'
  ) +
  scale_x_date(date_breaks = '3 months', date_labels = "%b %Y") +
  theme_classic() +
  theme(
    legend.position = 'top',
    axis.text.x = element_text(angle = 45, vjust = 1, hjust= 1)
  )
```
```{r}
#There is a cyclical weather pattern. From a glance it looks like there are more heating days then cooling days. 
```

```{r}
ggplot(daily_temp_Chicago[temp_series == 'avg_temperature_degrees_fahrenheit'],
aes(degrees, color = factor(year))) +
geom_density() +
labs(color ='', x = 'Average Temperature (Degrees Fahrenheit)',
y = 'Probability Density Function') +
theme_classic() +
theme(legend.position = 'top')

```
```{r}
#We notice the shape of the distribution to be skewed to the left. Out of the three years, 2022 looks like it has the most cold temperature days sub 30 degrees Fahrenheit. I believe the bimodal distribution comes from the the increase in probability of temperatures to be near 40-50 and then 75, based on the seasons. 

```

```{r}
ggplot(daily_temp_Chicago[temp_series == 'avg_temperature_degrees_fahrenheit'],
       aes(degrees, color = factor(year))) +
  geom_density() +
  labs(color = '', 
       x = 'Average Temperature (Degrees Fahrenheit)',
       y = 'Probability Density Function') +
  theme_classic() +
  theme(legend.position = 'top') +
  facet_wrap(~factor(month, levels = month.abb))
```
```{r}
#this lines up with our hypothesis of the bimodal model, since we see peaks in the winter seasons and summer seasons. Two higher averages would output that shape.
```

```{r}
lmp_temp_Chicago <- data_Chicago[,.(mean_lmp_rt = mean(total_lmp_rt),
    mean_lmp_da = mean(total_lmp_da),
    vol_spread = sd(spread),
    temp = mean(avg_temperature_degrees_fahrenheit),
    humidity = mean(relative_humidity_percent)),
  by = .(date, month, year)]

ggplot(lmp_temp_Chicago, aes(mean_lmp_rt, color = factor(year))) +
  stat_ecdf() +
  labs(x = 'Daily Average RT LMP ($/MWh)', y = 'Empirical Cumulative Distribution
Function',
  color = '') +
theme_classic() +
theme(legend.position = 'top')

ggplot(lmp_temp_Chicago, aes(temp, color = factor(year))) +
  stat_ecdf() +
  labs(x = 'Daily Average Temperature (Degrees Fahrenheit)', y = 'Empirical Cumulative
Distribution Function',
  color = '') +
  theme_classic() +
  theme(legend.position = 'top')

```
```{r}
#As stated above there is a clear FOSD with 2022 utilizing the highest daily average RT LMP. This is not consistent with the temperature graphs as we see 2023 is a warmer year than 2022. Temperature patterns may not have have translated to LMP patterns because there could have been more extreme volatile changes in certain days in 2022 (like in the winter), which drive LMP higher to be higher
```

```{r}
data_Chicago[, change_temp := shift(avg_temperature_degrees_fahrenheit, type = "lead") - avg_temperature_degrees_fahrenheit]



ggplot(data_Chicago, aes(x = month, y = change_temp, color = factor(year))) +
  geom_line() +
  labs(title = "Daily Percent Change in Temperature by Year",
       x = "Date",
       y = "Percent Change in Temperature (%)",
       color = "Year") +
  theme_classic()


```
```{r}
ggplot(data_Chicago, aes(x = date, y = change_temp, color = factor(year))) +
  geom_line() +
  labs(title = "Daily Percent Change in Temperature by Year",
       x = "Date",
       y = "Percent Change in Temperature (%)",
       color = "Year") +
  theme_classic()
```
```{r}
years <- c(2021, 2022, 2023)

for (y in years) {
  sd_val <- data_Chicago[year == y, sd(avg_temperature_degrees_fahrenheit)]
  print(paste("Year:", y, "SD:", sd_val))
}

```
```{r}
#We say the highest volatility in the average daily change in temp in 2022, which helps explain the increase in LMP. 
```


When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
